WGS Full Walkthrough

### 2022-02-18 WGS-Full-Walkthrough.Rmd
Update WGS high cov walkthrough

---
title: WGS fastq2vcf walkthrough

---

## Walkthrough Data
In this walkthrough, data files from whole genome sequences at 15x coverage from Atlantic cod will be referenced. Their naming struction are a population ID in the form of "Pop#" with the number being which population the sample came from, a "\_"\, and a 5 digit sample ID which is two digits for the year it was sampled and a three digit sample ID number. Here is an example, "Pop6_18001". We will walk through steps from demultiplexing when unique indexes are used through variant calling with Freebayes.

@@

#### Example Fastq File from Sequencing Facility 
To inspect an example data file, download files i5-4-i7-11_R1_exampleData.fastq.gz & i5-4-i7-11_R2_exampleData.fastq.gz which are truncated fastq files as examples for how data look from the sequencing facility.

@@

What these files show are XXXX.

### Stacks - process_radtags

Barcode File: For each set of paired end files that need to be demultiplexed, a barcode file needs to be created. For example, for files i5-4-i7-11_R1_exampleData.fastq.gz and i5-4-i7-11_R2_exampleData.fastq.gz its barcode file is as follows:

@@

This file contains the 12 barcodes that have the i5_4 and i7_11 index and which sample that combination belongs to. This file is tab delimited and because I had inline barcodes, meaning they were the same on both ends of the read, I have the same barcode sequence in the first and second column. Then the sample ID is in the far right column.

@@

We provide a simple R script to create the barcode files for each primer index combination. See file: Stacks_FileCreate.R

@@

```--inline_inline ``` The barcode option flag has 6 different options and you have to use the proper flag for your barcodes or indexes need to be demultiplexed. Because my samples have inline barcodes on either end of the sequence I need to use the --inline_inline flag.

```-P``` If paired end sequencing was used you need to add the -P flag to indicate this.

@@

```-r``` this flag allows Stacks to rescue any barcodes that are off by one bp

@@

#### Example code for running on a Cluster 
```
#!/bin/bash
#SBATCH --job-name=stacks_i54_i711                # sets the job name
#SBATCH --mem=1Gb                                 # reserves 1 GB memory
#SBATCH --mail-user=schaal.s@northeastern.edu     # add your email address to get notifications
#SBATCH --mail-type=FAIL                          # set what email notifications you want to receive here I only want emails when my script fails
#SBATCH --time=24:00:00                           # reserves machines/cores for 24 hours.
#SBATCH --output=stacksi54_i711.%j.out            # sets the standard output to be stored in file, where %j is the job id
#SBATCH --error=stacksi54_i711.%j.err             # sets the standard error to be stored in file

process_radtags -P -p i54_i711/GenomeFilesi54_i711/ -o ../Stacks_Out/ -b i54_i711/barcodeFilei54_i711.txt --inline_inline --disable_rad_check -r
```

### Results 
Time: 				each file took approx. 7 hrs  
Memory: 			approx. 80 MB RAM

@@

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop9_18101_submitFastP.txt
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --partition=lotterhos
#SBATCH --time=1:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=FastP_Out/jobSum/clustOut/Pop9_18101.%j.out
#SBATCH --error=FastP_Out/jobSum/clustOut/Pop9_18101.%j.err

@@

}
```
![Output](images/Pop1_17291100XcoveragePlot10000.png)

<!-- <img src="./Pop1_17291100XcoveragePlot10000.png" width="500">  -->

@@

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}

setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/WGSFullWalkthrough.Rmd")

>>>>

---
layout: page
title: WGS Full Walkthrough
permalink: /WGSwalkthrough/
---

@@

#### Example Fastq File from Sequencing Facility

@@

### Stacks Walkthrough

#### process_radtags walkthrough

Barcode File: For each set of paired end files that need to be demultiplexed, a barcode file needs to be created. For example, for files i5-4-i7-11_R1_001.fastq.gz and i5-4-i7-11_R2_001.fastq.gz its barcode file is as follows:

@@

This file contains the 12 barcodes that have the i5_4 and i7_11 index and which sample that combination belongs to. This file is tab delimited and because I had inline barcodes, meaning they were the same on both ends of the read, I have the same barcode sequence in the first and second column. Then the sample ID is on the right.

@@

We provide a simple R script to create the barcode files for each primer index combination. See file: Stacks_FileCreate_OmicsWG.R. You can use the practice files provide to see how the code works or just input your own to make the files you need. This file may be modified to fit your needs and there are annotations where that would be the case.

@@

```--inline_inline ``` The barcode option flag has 6 different options and you have to use the proper flag for your barcodes or indexes that need to be demultiplexed. Because the example samples have inline barcodes on either end of the sequence if using the practice data you need to use the --inline_inline flag.

```-P``` If paired end sequencing was used, you need to add the -P flag to indicate this.

@@

```-r``` this flag allows Stacks to rescue any barcodes that are off by one bp. For example, the barcode ACTTGA might appear as CTTGA or TACTTGA. This should only be a small percentage of reads and will probably recover about 1% of reads.

@@

#### Example code for running on a Cluster  
You can run the practice data files using the format below while changing the path on your computer to fit your needs. 
```
#!/bin/bash
#SBATCH --job-name=stacks_i54_i711                # sets the job name
#SBATCH --mem=1Gb                                 # reserves 1 GB memory
#SBATCH --mail-user=schaal.s@northeastern.edu     # set your email to get notifications about your run
#SBATCH --mail-type=FAIL                          # if the job fails for any reason email the provided email
#SBATCH --time=24:00:00                           # reserves machines/cores for 24 hours.
#SBATCH --output=stacksi54_i711.%j.out            # sets the standard output to be stored in file, where %j is the job id
#SBATCH --error=stacksi54_i711.%j.err             # sets the standard error to be stored in file

process_radtags -P -p ./genomeFilesi54_i711/ -o ./Stacks_Out/ -b barcodeFilei54_i711.txt --inline_inline --disable_rad_check -r
```
This will create four new fastq files for each sample provided in the barcodeFile. Because the practice data is a subset of the full dataset, this will only fill files with reads that are included in the subset file. The process_radtags program wants to keep the reads in phase, so that the first read in the sample_XXX.1.fq file is the mate of the first read in the sample_XXX.2.fq file. Likewise for the second pair of reads being the second record in each of the two files and so on. When one read in a pair is discarded due to low quality or a missing restriction enzyme cut site, the remaining read can't simply be output to the sample_XXX.1.fq or sample_XXX.2.fq files as it would cause the remaining reads to fall out of phase. Instead, this read is considered a remainder read and is output into the sample_XXX.rem.1.fq file if the paired-end was discarded, or the sample_XXX.rem.2.fq file if the single-end was discarded.

There will also be a log file that summarizes the number of reads per sample that were retained, were low quality, or that had noRADtag (only relevant for rad data).

#### Results for a run with one primer combination on the full paired end data files that for each ~30GB in size
Time: 				each file took approx. 7 hrs  
Memory: 			approx. 80 MB RAM
Run description: run on separate cluster nodes for 32 different primer combinations

@@

#### Example for running on a Cluster
```
#!/bin/bash
#SBATCH --job-name=Pop9_18101_submitFastP.txt
#SBATCH --mem=2Gb
#SBATCH --mail-user=schaal.s@northeastern.edu
#SBATCH --mail-type=FAIL
#SBATCH --time=1:00:00
#SBATCH --nodes=1
#SBATCH --tasks-per-node=1
#SBATCH --output=FastP_Out/jobSum/clustOut/Pop9_18101.%j.out
#SBATCH --error=FastP_Out/jobSum/clustOut/Pop9_18101.%j.err

@@

}
```
<img src="./Pop1_17291100XcoveragePlot10000.png" width="500">

@@

```{r, echo=FALSE,eval=FALSE}
library(knitr)
my.jekyll.site <- "marineomics.github.io"
KnitPost <- function(input, base.url = my.jekyll.site) {
  opts_knit$set(base.url = base.url)
  fig.path <- paste0("images/", sub(".Rmd$", "", basename(input)), "/")
  opts_chunk$set(fig.path = fig.path)
  opts_chunk$set(fig.cap = "center")
  render_jekyll()
  knit(input, envir = parent.frame())
}
setwd("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines")
KnitPost("~/Postdoc_Projects/RCN_MarineOmics/Website/marineomics.github.io/guidelines/WGSFullWalkthrough.Rmd")
```
